Lossy compression of DNA sequencing quality data

### 2015-02-03 manuscript/q-compression.tex
More revisions, particularly to the intro, new references, other changes.

Sequencing read data constitutes the largest category of sequence data
in genomics research because of great redundancy, inclusion of quality
values, and read-level naming and metadata. Because of this abundance
effective compression of read data has the potential for substantial
improvement in data storage and transfer efficiency.  Among the
several challenges for compression of read data is dealing with
different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies. Illumina sequencing reads, the focus of this
research, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low quality tails at the 3$^{\prime}$
end~\cite{Minoche:2011km}).

Quality values comprise a standard component of \textsc{fastq} format
files~\cite{Cock:2010ve}.  At the level of sequence read the
probability of error for each base-call is typically represented by
\textsc{phred} quality value, which is defined as $Q =
-10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
technology these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). These errors are represented a single quality value per
base-call for Illumina sequencing reads.

Quality values are typically used throughout bioinformatics pipelines.
Software such as \textsc{fastqc}~\cite{citeulike:11583827} and
Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
descriptions of the distribution of sequence quality scores, and these
scores can be used to guide filtering and trimming of sequencing read
data as processing step before subsequent analyses. Quality values can
be used by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Gnerre:2011kx}.  Short-read alignment
software, such as Bowtie 2~\cite{Langmead:2012rw}, use quality values
to scale the mismatch penalty in a way that mismatches between poor
quality bases of the query and the reference are weighted less than
mismatches occurring on high quality bases. Similar to alignment,
software for detecting single nucleotide polymorphisms (\textsc{snp}s)
can use quality values~\cite[e.g.,][]{McKenna:2010bh}, and identified
\textsc{snp}s with high-quality calls are deemed more reliable than
those with low-quality calls, particularly in low coverage regions.

Although the quality values are useful for later steps, the trade-off
between storing effectively double the amount of data and the
resulting effects on subsequent analyses has not been
well-documented. Previous literature on sequence data compression has
largely focused on lossless compression of base
calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
Typical compression methods fail to utilize the biases inherent in
existing sequencing technologies. Depending on the sequencing
technology used, the base calling is more accurate in the earlier
cycles, gradually losing quality in later
cycles~\cite{Kozanitis:2011kl}. To this end, methods for compressing
these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

Little prior work has been done examining the effect of lossy
compression of quality sequences in the context of subsequent
analyses. Our aim here is to examine the effect of multiple lossy
compression methods on quality sequences, focusing not only on the
differences between the compressed and original versions, but also the
effect of lossy compression on typical subsequent analyses.

We examine three general classes of lossy compression methods:
binning, modeling, and profiling.

@@

\subsection{Lossy compression effect on preprocessing quality}

@@

\subsection{Lossy compression effect on assembly quality}

@@

\begin{table*}[!tpb]
\centering
\caption[]{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
  2.4.1. data set provided by
  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
  by their overall rank (determined by the sum of individual
  rankings).}
{\begin{tabular}{lcccccccc}
& Overall & & & & & & \textsc{lap} & \textsc{lap} \\
Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
\hline
None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
Profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
Regression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
2-bin	               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
Regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
Regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
Regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

\subsection{Lossy compression effect on mapping quality}

@@

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{llcccc}
    & & \multicolumn{4}{c}{Percentage of reads mapped}\\
    \cline{3-6}
	Library & Compression Strategy & Total & Shared & Uncompressed only & Compressed only \\ \hline
	& Maximum quality & 77.88 & 77.88 & 22.12 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\
	&2-bin & 99.92 & 99.89 & 0.11 & 0.03 \\
	&Regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
	&Regression (1st order) & 98.62 & 98.61 & 1.39 & 0.01 \\
	&Regression (2nd order) & 99.42 & 99.39 & 0.61 & 0.03 \\
	Fragment&Regression (3rd order) & 99.56 & 99.53 & 0.47 & 0.03 \\
	&Regression (4th order) & 99.71 & 99.68 & 0.32 & 0.03 \\ 
	&Regression (5th order) & 99.79 & 99.75 & 0.25 & 0.03 \\
	&Profile (32) & 99.9 & 99.86 & 0.14 & 0.04 \\ 
	&Profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\

@@

	&Maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\ 
	&2-bin & 97.38 & 97.38 & 2.62 & 0 \\
	&Regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
	&Regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
	&Regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
	Short-jump&Regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
	&Regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
	&Regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
	&Profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
	&Profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
\end{tabular}

@@

\begin{table}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the
  proportion of bases/reads that are missed or gained after applying
  compression with respect to the uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lcccc}
	 & \multicolumn{2}{c}{Fragments} & \multicolumn{2}{c}{Short-Jump} \\
	 & \% Bases & \% Reads & \% Bases & \% Reads \\ \hline
	2-bin & 98.15 & 98.78 & 113.09 & 103.7 \\ 
	Regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
	Regression (1st order) & 92.96 & 95.09 & 81.31 & 91.79 \\ 
	Regression (2nd order) & 93.93 & 96.87 & 86.31 & 94.75 \\
	Regression (3rd order) & 93.5 & 95.68 & 88.7 & 94.51 \\ 
	Regression (4th order) & 94.43 & 96.7 & 91.94 & 97.02 \\ 
	Regression (5th order) & 95.56 & 98.09 & 93.92 & 98.02 \\ 
	Profile (32) & 99.12 & 98.96 & 102.89 & 102.24 \\
	Profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
\end{tabular}
\end{table}

>>>>

Sequencing read data constitutes the largest category of sequence data
in genomics research because of great redundancy, inclusion of quality
values, and read-level naming and metadata. Because of this abundance
effective compression of read data has the potential for substantial
improvement in data storage and transfer efficiency.  Among the
several challenges for compression of read data is dealing with
different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies. Illumina sequence reads, the focus of this
research, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low quality tails at the 3$^{\prime}$
end~\cite{Minoche:2011km}).

Quality values comprise a standard component of \textsc{fastq} format
files~\cite{Cock:2010ve}.  At the level of sequence read the
probability of error for each base-call is typically represented by
\textsc{phred} quality value, which is defined as $Q =
-10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
technology these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). These errors are represented a single quality value per
base-call for Illumina sequence reads.

Quality values are typically used throughout bioinformatics pipelines.
Software such as \textsc{fastqc}~\cite{citeulike:11583827} and
Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
descriptions of the distribution of sequence quality scores, and these
scores can be used to guide filtering and trimming of sequence read
data as processing step before subsequent analyses. Quality values can
be used by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Gnerre:2011kx}.  Short-read alignment
software, such as Bowtie 2~\cite{Langmead:2012rw}, use quality values
to scale the mismatch penalty in a way that mismatches between poor
quality bases of the query and the reference are weighted less than
mismatches occurring on high quality bases. Similar to alignment,
software for detecting single nucleotide polymorphisms (\textsc{snp}s)
can use quality values~\cite[e.g.,][]{McKenna:2010bh}, and identified
\textsc{snp}s with high-quality calls are deemed more reliable than
those with low-quality calls, particularly in low coverage regions.

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
Although some recent work has included or focused on compression of
quality
values~\cite[e.g.,][]{Tembe:2010ys,Kozanitis:2011kl,asnani2012lossy,Hach:2012ys,Wan:2012kq,janin2013adaptive,Ochoa:2013rt,Canovas:2014fr,zhou2014compression,DBLP:conf/recomb/YuYB14}.
Typical compression methods fail to utilize the biases inherent in
existing sequencing technologies. Depending on the sequencing
technology used, the base calling is more accurate in the earlier
cycles, gradually losing quality in later
cycles~\cite{Kozanitis:2011kl}. To this end, methods for compressing
these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some absolute loss of
information is deemed acceptable if the inferences from analyses are
relatively uneffected. Therefore we here describe our research
investigating some general strategies for compression of sequence read
quality values. Our objective provide perspective on several
strategies logically applied to Illumina sequence read quality values
rather than to generate a robust high-quality software for
use. Consistent with our perspective we place emphasis not only on the
compression efficiency, but also the effect of lossy compression on
typical subsequent analyses.  We examine three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
consider an exemplar of each class.

@@

\subsection{Effects on sequence read preprocessing}

@@

\subsection{Effects on genome assembly}

@@

\begin{table*}[!tpb]
\centering
\caption[]{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
  2.4.1. data set provided by
  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
  by their overall rank (determined by the sum of individual
  rankings).}
{\begin{tabular}{lcccccccc}
& overall & & & & & & \textsc{lap} & \textsc{lap} \\
compression            & rank & \textsc{snp}s & inversions & relocations & translocations & indels & fragment & short-jump \\
\hline
none                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
pegression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
2-bin	               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

\subsection{Effects on read mapping}

@@

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{llcccc}
    & & \multicolumn{4}{c}{percentage of reads mapped}\\
    \cline{3-6}
	library & compression strategy & total & shared & uncompressed only & compressed only \\ \hline
	& maximum quality & 77.88 & 77.88 & 22.12 & 0 \\
	&minimum quality & 100 & 100 & 0 & 0 \\
	&2-bin & 99.92 & 99.89 & 0.11 & 0.03 \\
	&regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
	&regression (1st order) & 98.62 & 98.61 & 1.39 & 0.01 \\
	&regression (2nd order) & 99.42 & 99.39 & 0.61 & 0.03 \\
	fragment&regression (3rd order) & 99.56 & 99.53 & 0.47 & 0.03 \\
	&regression (4th order) & 99.71 & 99.68 & 0.32 & 0.03 \\ 
	&regression (5th order) & 99.79 & 99.75 & 0.25 & 0.03 \\
	&profile (32) & 99.9 & 99.86 & 0.14 & 0.04 \\ 
	&profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\

@@

	&maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
	&minimum quality & 100 & 100 & 0 & 0 \\ 
	&2-bin & 97.38 & 97.38 & 2.62 & 0 \\
	&regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
	&regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
	&regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
	short-jump&regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
	&regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
	&regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
	&profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
	&profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
\end{tabular}

@@

\begin{table}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the
  proportion of bases/reads that are missed or gained after applying
  compression with respect to the uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lcccc}
	 & \multicolumn{2}{c}{fragments} & \multicolumn{2}{c}{short-Jump} \\
	 & \% bases & \% reads & \% bases & \% reads \\ \hline
	2-bin & 98.15 & 98.78 & 113.09 & 103.7 \\ 
	regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
	regression (1st order) & 92.96 & 95.09 & 81.31 & 91.79 \\ 
	regression (2nd order) & 93.93 & 96.87 & 86.31 & 94.75 \\
	regression (3rd order) & 93.5 & 95.68 & 88.7 & 94.51 \\ 
	regression (4th order) & 94.43 & 96.7 & 91.94 & 97.02 \\ 
	regression (5th order) & 95.56 & 98.09 & 93.92 & 98.02 \\ 
	profile (32) & 99.12 & 98.96 & 102.89 & 102.24 \\
	profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
\end{tabular}
\end{table}
